### 2020-08-06 paper/sex_selection_spacing-ver4.tex
Editing conclusion

% For word counting uncomment line below and raggedright below
% http://countwordsfree.com/#
\usepackage[nolists]{endfloat}

@@

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth intervals. 
Over the four decades, the median birth interval for this group increased by almost 15, 
and the 75th percentile birth interval increased by a staggering 21 months, most of that 
within a decade.

@@

% Background
Over the last four decades India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained. 
At the same time, economic growth was strong, schooling increased, and the total
fertility rate fell to close to replacement.

% Questions
The main question I address in this paper is how birth spacing responded to the major changes
that occurred in India over the last four decade, in particular the spread of sex selection.
In addition, I examine two related questions.
First, did the changes in birth spacing bias the standard fertility estimates for India? 
Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?

% What has happened to spacing and why
Birth intervals has lengthen, often substantially, over the last four decades.
The increases in spacing are larger, the higher the parity and the higher the mother's
education, and is larger in urban than in rural areas.

Groups with the highest use of sex selection have the most substantial increases in 
birth intervals, with increases in average birth intervals of up to 12 months. 
Some of these increases are so large that we even observe a reversal of the traditional 
spacing pattern: instead of the absence of sons leading to the shortest birth intervals, 
son preference and sex selection lead some groups to have longest birth intervals when
they have no sons.

Despite the longer average birth intervals, a large number of women still have very 
short birth intervals. 
Hence, the changes in birth spacing are driven predominately by the longer intervals 
becoming even longer over time.
This pattern holds generally, but is especially prevalent for those who use sex selection 
and, therefore, may end up with multiple abortions.

% fertility
Fertility has declined for all groups, but the likelihood of having an additional child
still depends strongly on the number of sons, with women with no sons having the highest 
parity progression probabilities.

The combination of the secular increases in spacing and the fact those women most likely to use
sex selection are also those most likely to have a next birth, makes the total fertility rate a 
more biased measure of cohort fertility. 
This bias was most prominent early in the spread of sex selection, when the fertility rate 
was up to one child lower than the predicted cohort fertility. 
It is, however, still present, with the predicted cohort fertility ten to twenty percent 
higher than the fertility rate.

Tempo effects are studied extensively in the literature \citep[See, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in the total fertility rate as birth spacing 
stabilizes or, if interventions against sex selection are successful, even shortens again.

The best-educated urban women are the only group for whom the predicted cohort fertility 
is below replacement at 1.8. 
A caveat to the predicted cohort fertility is that only Hindu women are in the sample and 
that the model only includes births up to parity four.
However, almost everybody marries, and Hindu constitute 80 percent of the population.
Furthermore, with only the first fourth births used, these estimates may
be lower-bound, especially for the less educated women.

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the difference is 
small for the best-educated women.
There is no evidence that children born to women who are the most likely to use sex 
selection have higher mortality risk.

An important implication of the results here is that population growth in India
may be higher than expected. 
There has been a rapid reduction in infant mortality risk, the longer spacing should 
further reduce mortality risk, and cohort fertility is higher than previously realized.

@@

There are two critical questions that future research should address.
First, sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an
unintended side-effect, rather than a result of girls becoming more
valued as is often assumed. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.

Second, is there a relationship between female labor force participation
and sex selection? 
Women may be staying out of the labor market precisely because sex selection makes 
them more likely to have a boy and increases the expected birth spacing. 
Better job opportunities for women would affect sex selection for two reasons. 
First, it makes it more expensive to be out of the labor market for long periods. 
Second, it would moderate the differential in potential earnings between husband
and wife and make it more attractive to invest in female compared to
male offspring's human capital. 
This approach could, however, be a double-edged sword. 
If better job opportunities further lower desired fertility, the use of sex selection 
may increase, everything else equal.
Hence, understanding the trade-off between long-term benefits from
improvements in women's labor force participation and short-term costs
from potential increases in sex selection is of paramount importance.

@@

[More formal version of empirical strategy section]

>>>>

% For word counting uncomment line below and raggedright below
% http://countwordsfree.com/#
% \usepackage[nolists]{endfloat}

@@

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth intervals. 
Over the four decades, the median birth interval for this group increased by almost 15 
months, 
and the 75th percentile birth interval increased by a staggering 21 months, most of that 
within a decade.

@@

% Background
Over the last four decades, India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained high. 
Simultaneously, economic growth was strong, schooling increased, and the total
fertility rate fell to close to replacement.

% Questions
The main question I address in this paper is how birth spacing responded to the 
significant changes in India over the last four decades, particularly the spread of sex 
selection.
I also examine two related questions.
First, did the changes in birth spacing bias the standard fertility estimates for India? 
Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women
because of their substantial use of sex selection combined with falling fertility.
As the sex ratio increased to almost 80 percent boys, the expected median birth interval 
for a woman with twelve or more years of education who has two girls increased by almost 
15 months, and the 75th percentile interval by 21 months.
Most of that increase came right after the introduction of sex selection in India.

Some of these increases are so large that we even observed a reversal of the traditional 
spacing pattern for some groups; when there are no sons, we now see the longest, rather 
than the shortest, birth intervals because of sex selection.
The women who are the least likely to use sex selection still show the traditional 
spacing pattern with short spacing in the absence of sons.

Son preference continues to show in fertility decisions. 
Fertility has declined for all groups, but the likelihood of having an additional child
still depends strongly on the number of sons, with women with no sons having the highest 
parity progression probabilities.

Birth intervals also lengthened in cases when sex selection is less likely to be used. 
However, compared to other countries with similar declines in fertility, the median 
spacing increases were smaller at three to six months over the four decades. 
Most of the median intervals when sex selection is less used are still short at 36 months 
or below. 
Furthermore, many women still have very short birth intervals. 
In many cases, more than 25 percent have their next child within 24 months of the previous 
birth.

% [predictions of declining sex selection] 
Despite predictions that the use of sex selection would decline, there is no clear 
evidence of this. 
The original users of sex selection continue to show substantial male-biased sex ratios, 
although there may be some leveling off. 
More concerning, sex selection appears to be spreading to less educated women as their 
fertility is falling.

% fertility bias

The increases in spacing make the total fertility rate a more biased measure of cohort 
fertility. 
This bias was most prominent early in the spread of sex selection when the fertility rate 
was up to one child lower than the predicted cohort fertility. 
However, it is still present, with the predicted cohort fertility ten to twenty percent 
higher than the fertility rate. 
At 1.8 children, the best-educated urban women are the only group for whom the predicted 
cohort fertility is below replacement.

Tempo effects are studied extensively in the literature \citep[See, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in the total fertility rate as birth spacing 
stabilizes or even shortens again if interventions against sex selection are successful.

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that substantial use of sex selection is associated with higher 
infant mortality.

% Implications 
An important implication of the results here is that population growth in India may
not have fallen as much as recently suggested.
There has been a rapid reduction in infant mortality risk, the longer birth spacing should 
further reduce mortality risk, and cohort fertility is higher than previously realized.

@@

There are two critical questions that future research should address.
First, sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an unintended 
side-effect, rather than the result of girls becoming more valued as is often assumed. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.

Second, what is the relationship between female labor force participation and sex selection? 
Women may be staying out of the labor market precisely because sex selection makes 
them more likely to have a boy and increases the expected birth spacing. 
Better job opportunities for women would affect sex selection for two reasons. 
First, it makes it more expensive to be out of the labor market for long periods. 
Second, it would moderate the differential in potential earnings between husband and wife 
and make it more attractive to invest in daughters' human capital. 
This approach could, however, be a double-edged sword. 
If better job opportunities further lower fertility, the use of sex selection may increase, 
everything else equal.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.

@@

% [More formal version of empirical strategy section]
